# DATA IMPORT
<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

<script src="files/js/dcl.js"></script>
```{r ,include=FALSE}
tutorial::go_interactive(greedy=TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```

## THE WORKING DIRECTORY

In R, a working directory is the default location where R looks for
files you load, save and/or access without providing a full path. It is
like your home base for all file operations within R.

R uses file paths to locate files on your computer's file system.

In R, you can use an absolute or relative path to specify a working
directory. If you are a bit familiar with bash, a relative path is the
path relative to your working directory. For example if you set your
working directory to a folder called `project`, to load a dataset called
`quarterly_sales.csv` in a subfolder called `data` for example, the
relative path would be `project/data/quarterly_sales.csv`.

The absolute path could be something like this:
`C:/Users/YourName/Documents/project/data/quarterly_sales.csv`

Working in RStudio, file systems can be a bit of a nuisance if you do
not correctly set them up. This is why it is recommended to create a
project in RStudio whenever you're working with it as it will minimize
your file system headaches a lot.

We have already talked about setting up a project in Chapter 1 but,
sometimes in R you might want to work without setting up an entire
project. In this case, you need to set up a working directory. Why?
Setting up a working directory keeps your files organized, ensures
portability and overall, makes the overall development experience
convenient.

### How to setup a working Directory

In RStudio, there are three main ways to set up a working directory. You
can use the Menu bar, a keyboard shortcut or you can set it
programatically.

#### RStudio Menu

To set up a working directory, go to the menu bar at the top of your
RSTudio window.

-   Go to "Session" \> "Set Working Directory" \> "Choose Working
    Directory"
-   Select your desired directory and click "Choose"

#### Keyboard Shortcut

-   For Windows & Linux: Press Ctrl+Shift+H
-   For macOS: Press Cmd+Shift+H

#### Setting Working Directory Programmatically

To do this, you can use base R's `setwd()` function and supply the file
path, whether relative or absolute.

Here is an example:

```{r,tut=TRUE,height=500}
# Set working directory
#setwd("/path/to/your/desired/directory")
```

Where `/path/to/your/desired/directory` is the actual path you want to
set it to.

## IMPORTING DATA INTO R

Importing data into R is the process of bringing in data from external
sources into R's environment for further analysis ot manipulation. It is
usually the first step in almost every R workflow.

### Data Sources

R's prowess is such that, you can import nearly any data file format.
From tabular .txt files, .csv, delimited text files, SAS, SPSS, Stata,
Excel, JSON and APIs as well as .jpeg, .tiff, Big data and even
importing data from databases, R does it all. We talked about the
`read.csv()` function in the previous chapter but that isn'T all. There
are numerous functions from various packages to read in data into R,
however, there are some default functions that do this. The built-in
functions for data import into R are: - `read.table()`: Reads data from
a tabular text file. - `read.csv()`: Reads data from a CSV file. -
`read.delim()`: Reads data from a delimited text file. - `readLines()`:
Reads lines from a text file.

### The Importing Process

Importing data into R is quite simple as functions follow similar logic.
You will need a file path, the appropriate function and it is usually
recommended to assign the imported data to a variable to prevent it from
being printed and lost.

Following this workflow, we can import a `.csv` file like this:

```{r,tut=TRUE,height=500}
# Syntax for importing a dataset
#my_data <- read.csv("data/my_data.csv")
```

The line has been commented out because it is used to show the syntax of
importing data into R. You will notice we wrapped the file path of the
dataset including our dataset name and extension in quotes. This is
important as failure to do so will make R interpret is as a variable and
throw up an error. We then assigned our imported data to a variable
called `my_data`.

### Data Exploration

After importing data, R has some built-in functions you can use to
inspect your dataframe's contents and structure.

You can use the following functions to achieve this: - `View()`- opens a
spreadsheet-like viewer to explore dataframe interactively - `head()` -
displays first few rows of your dataframe - `tail()` - just like head,
it displays last few rows of your dataframe - `str()` - stands for
structure and it is used to display a compact internal structure of an R
object. It usually shows each type of variable in your dataframe, its
length and a few values of each variable.

## DATA FROM SOME STATISTICAL PACKAGES INTO R

There are a lot of packages that allow imports from data sources into R.
To use them, you usually need to install and load them into your R
environment before accessing these import functions.

Here are some common Data Sources and R Functions to import them with.
It is recommended to read the package documentation for specific options
or parameters.

-   Text files (CSV, TSV): `read.csv()`: For comma-separated values
    `read.table()`: For more general text files, with options to specify
    delimiters
-   Excel files: `readxl::read_excel()`: Requires the readxl package
-   JSON files: `jsonlite::fromJSON()`: Requires the jsonlite package
-   SQL databases: `RSQLite::dbConnect()`: For SQLite databases
    `DBI::dbConnect()`: For other database types (requires specific
    database drivers)
-   Statistical packages (SPSS, SAS): `foreign::read.spss()`: For SPSS
    files `haven::read_sav()`: For SAS files (requires the haven
    package)
-   Web APIs: `httr::GET()`: To retrieve data from web APIs
    `jsonlite::fromJSON()`: To parse JSON responses

## Data EXPORT

Just like we can import data into R, R also allows us to export data
from R to our local disk or wherever we want to export to.

For text files, you can use R's built-in `write.csv()` to write a
dataframe to a `.csv` file or `write.table()` for text files. you can
also use `saveRDS()` to save an R object in R readable format or
`save()` to save multiple objects to an `RData` file.

For other formats like excel files you can use the package `writexl` to
do that by calling `write.xlsx()` to save a dataframe to an excel sheet.

For JSON files, you can use the function `toJSON()` from `jsonlite` to
write a file to JSON format.

### Writing to different formats
To export to as a `.csv` file you use the 
```{r,tut=TRUE,height=500}
# Create a sample dataframe
df <- data.frame(
  ID = c(1, 2, 3, 4, 5),
  Name = c("Thomas", "Bob", "Helen", "David", "Eva"),
  Age = c(25, 30, 22, 35, 28),
  Score = c(95, 85, 92, 88, 75)
)

# You can save to your project directory like this. Lines have been commented
# out because we do not have 
# write.csv(df, file = "output_data.csv", row.names = FALSE)
# write.table(df, file = "output_data.txt", sep = "\t", row.names = FALSE)
# write_xlsx(df, path = "output_data.xlsx")

# For JSON, you need to convert the data to JSON format before exporting. 
# json_data <- toJSON(df, pretty = TRUE)
# writeLines(json_data, "output_data.json")
```

For JSON objects, it is always recommended to ensure your dataframe is
compatible with a JSON format before exporting.

## OTHER DATA FORMATS

There are countless other data formats R supports. Usually, as an R
programmer, you will have to search for these packages or if you have
the skill and there is enough interest, build these packages yourself as
R extensions to support data types you want.

### Web Scraping

An Aspect we would like to briefly touch on is web scraping. Sometimes
the data we want are available on websites and not in necessarily ready
formats to allow further analysis. This is where web scraping comes in.
R's prowess extends to this domain and allows data from websites to be
programmatically extracted into R for further analysis.

There are several web scraping packages but the most widely used ones in
R are the `httr` and `rvest` packages.


## ASSIGNMENT
- Create a project named `pilot` in RStudio
- Use the `datasets` package to import the `chickwts` dataset and assign to
  a variable called `df`
- Filter df for only `horsebean` fed chicks.
- Use `write.csv()` to export your dataset to your working directory
- Run the `dir()` function and copy its output to a `.txt` and submit it.
