# DESCRIPTIVE STATISTICS

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

<script src="files/js/dcl.js"></script>
```{r ,include=FALSE}
tutorial::go_interactive(greedy=TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```

Descriptive statistics in R involve the use of various numerical and graphical
methods to summarize and describe the main features of a dataset. These
statistics provide a concise overview of the central tendency, dispersion, and
shape of the distribution of data. Common measures include the mean, median,
and mode for central tendency, while measures such as range, variance, and
standard deviation describe the spread or dispersion of the data. Additionally,
summary statistics like minimum and maximum values, quartiles, and percentiles
help understand the overall distribution. R provides a rich set of functions,
including `summary()`, `mean()`, `median()`, `sd()` (standard deviation), and
various quantile functions, to quickly compute and explore these descriptive
statistics for a given dataset.

Visualization is also a key aspect of descriptive statistics in R. Box plots,
histograms, and density plots can be generated to visually represent the
distribution of data. Exploring descriptive statistics aids in uncovering
patterns, identifying outliers, and gaining a fundamental understanding of the
dataset's characteristics, setting the stage for more advanced statistical
analyses.


## MEASURES OF CENTRAL TENDENCY
Measures of central tendency, also called, measures of central location are
numerical values that summarize a dataset by indicating its central or typical
value.

The measures of central tendency in R are mainly the mean, the median and the
mode.

- The mean is the average value if you were to distribute the total evenly 
  among all observations
- The Median represents the middle value with half of the observations below
  and half above it
- The Mode is used to identify the most common value, which may highlight
  potential trends or patterns.

In R, using the `summary()` function summarizes a dataframe. It will return
the mean, median, quartiles, minimum and maximum values for each
numeric/integer column in your dataset along with indicating if there are
missing values in said columns.

Let us see the `summary()` function in action.

For this chapter, we will use the `mtcars` dataset from R's `datasets package`

```{r,tut=TRUE,height=500}
# Import the mtcars dataset
df <- datasets::mtcars

# Summarize the dataframe using summary
summary(df)
```

Simple as that! R summarizes the numeric and integer columns and gives us some
descriptive statistics which helps us understand our data before further
analysis.

### The mean
The mean is a measure of central tendency that represents the average value
of a set of numbers. It is calculated by adding up all the values in the
dataset and then dividing the sum by the total number of values. In simpler
terms, it's the sum of all values divided by the count of values.

In R, you can calculate the mean of a set of numbers like this:

```{r,tut=TRUE,height=250}
# Create a numeric vector
numeric_vector <- c(10, 15, 20, 25, 30)

# Calculate the mean using the mean() function
mean_value <- mean(numeric_vector)

# Print the result
mean_value
```

#### Handling Missing Values
The `mean()` function in R has an argument called `na.rm` (NA remove) that is
set to `FALSE` by default. This means that if there are any missing values
(`NA`s) in your data, the result of running the mean function will be `NA`. 

If you want to exclude missing values from the calculation, you can set
`na.rm = TRUE`. Here is an example

```{r,tut=TRUE,height=250}
# Create a vector with missing values
vector_with_na <- c(10, 15, NA, 25, 30)

# Calculate the mean, excluding missing values
mean_value_without_na <- mean(vector_with_na, na.rm = TRUE)

# Print the result
mean_value_without_na
```

### The median
The median is the middle value in a sorted set of values. To calculate the
median in R, you can use the `median()` function.

```{r,tut=TRUE,height=250}
# Example with a numeric vector
numeric_vector <- c(1, 2, 3, 4, 5)

median_value <- median(numeric_vector)

print(median_value)
```

You can also find the median of a numeric or integer column of a dataframe like
this:

```{r,tut=TRUE,height=250}
# Example with a data frame
data <- data.frame(Name = c("Alice", "Bob", "Brad", "Charlie", "Charlotte",
                            "David", "Daniela"),
                   Age = c(25, 50, 44, 20, 22, 28, 17))

median_age <- median(data$Age)

print(median_age)

```

Load the mtcars dataset and find the median cylinder value as well as the mean
horsepower in that dataset below

```{r,tut=TRUE,height=500}
# Write your code below

```



## MEASURES OF VARIABILITY
Also known as measures of dispersion, measures of variability tells us how
spread out the data is in a dataset. In other words, these statistics quantify
how much individual data points in a distribution deviate from the central
tendency. 

This information is crucial for understanding our data's nature and helps in
drawing accurate conclusions from our analysis


The key measures of variability in R are
- Range
- Interquartile Range (IQR)
- Variance
- Standard Deviation

### Range
Range is the simplest measure of variability.  It is usually calculated as the
difference between the largest and smallest values in a dataset. Range is
relatively easy to understand and straightforward but a huge issue with it is
that, it is very sensitive to outliers and does not tell much about the
distribution of the data between.

In R, we can use the min and max functions to find the highest and lowest
values but R has a built-in `range()` function that returns the range of 
given distribution

For example, let us use a vector to illustrate how this works. Do note that
you can also use the range function on a column in a dataframe but we will
use a vector here for simple illustration.

```{r,tut=TRUE,height=250}
# Generate a vector
data <- c(10, 5, 8, 12, 15, 8, 8)

# Call the range function
range(data)
```

The range of this given distribution is 5, the lowest value and 15, the
highest value.

Can you calculate the range of bill lengths in the penguins dataset below?

```{r,tut=TRUE,height=500}
# Import penguins dataset
#penguins <- read.csv('')

# Call the range function on the bill length column.
# Do you remember how to reference a dataframe column?


```

### Interquartile Range (IQR)
This divides our data into 4 quartiles, with the IQR representing the range
between the first and third quartiles (Q1 and Q3).

The interquartile range is less sensitive to outliers than the range and
provides information about the central 50% of a data's distribution.

```{r,tut=TRUE,height=500}
# Generate a vector
data <- c(10, 5, 8, 12, 15, 8, 8)

# Call the IQR function
IQR(data)

# Call the quantile function
quantile(data)
```

### Variance
Variance is the average squared distance of each data point from the mean. To
calculate it, we use the `var()` function.

```{r,tut=TRUE,height=250}
# Generate a vector
data <- c(10, 5, 8, 12, 15, 8, 8)

# Call var function
var(data)
```

### Standard Deviation
Standard deviation is the square root of the variance which represents the
typical distance of data points from the mean in the units of the original
data.

To calculate this we use the `sd()` function in R

```{r,tut=TRUE,height=250}
# Generate a vector
data <- c(10, 5, 8, 12, 15, 8, 8)

# Find standard deviation
sd(data)
```

Find the standard deviation of the body mass of the penguins in the 
palmerpenguins dataset

```{r,tut=TRUE,height=500}
# Import penguins dataset
#penguins <- read.csv('')

# Call the standard deviation function on the bill length column.
# Do you remember how tor reference a dataframe column?


```


You might be wondering why these functions work on vectors but also on
dataframe columns like above. The reason is that, these functions are built to
operate seamlessly on both individual vectors and dataframe columns.

They are perfect for vectors, but when you reference a specific column in a
dataframe and apply these functions, R treats that column as a vector in the
background. This underlying treatment as a vector allows the functions to work
effortlessly on dataframe columns as well.


### Factors for Choosing an Appropriate Measure
- Shape of distribution: for normal distributions, all measures correlate well.
  For skewed distributions, IQR and SD may be more informative.
- Presence of outliers: IQR and SD are more robust to outliers than range and
  variance.
- Interpretability: SD is easier to understand in the context of the original
  data units.


## SKEWNESS AND KURTOSIS
Skewness and kurtosis are two important quantitative measures used in
statistics and data analysis to describe the shape of a distribution. They 
provide crucial information beyond just the central
tendency (e.g., mean, median) of a dataset.


### Skewness
Skewness measures the asymmetry of a probability distribution. It indicates
whether the data is skewed to the left or right of the mean. Skewness can be
positive, negative or zero. 

When a distribution is said to be positively skewed, a visualization of the
distribution of its values will have a long tail to the right, indicating the
right side has few but larger values. In a positively skewed distribution, the
mean is typically greater than the median. 

When a distribution is said to be negatively skewed however, it has a long
tail to the left, indicating few but smaller values. The mean is typically
less than the median in a  negatively skewed distribution.

A distribution is considered symmetric if it has zero skewness, that is, the
tails on both sides of the mean are of equal length and the distribution
appears roughly symmetrical.


In R, you can calculate skewness using the skewness() function from the `e1071`
package or the skewness() function from the moments package. Here's an example
using the `e1071` package:

```{r,tut=TRUE,height=500}
# Using the e1071 package
#install.packages("e1071")

# Load the e1071 package
library(e1071)

# Create a numeric vector (replace this with your own data)
data <- c(10, 5, 8, 12, 15, 8, 8)

# Calculate skewness
skewness_value <- skewness(data)

# Print the skewness value
print(skewness_value)
```

We can conclude that our data is positively skewed.


### Kurtosis
Kurtosis is a statistical measure that describes the distribution of data in
terms of the tails and the peakedness or flatness of the distribution. It
provides insights into whether the data has heavy tails or is more concentrated
around the mean compared to a normal distribution. Kurtosis complements
skewness, which measures the asymmetry of the distribution.

There are three main types of kurtosis:

1. Mesokurtic (Normal Distribution):
  A distribution with kurtosis equal to zero is considered mesokurtic.
  It has tails and a peak similar to the normal distribution.

2. Leptokurtic (Positive Kurtosis):
  A distribution with positive kurtosis has heavier tails and a more peaked
  center compared to a normal distribution.
  It indicates that the data has more extreme values, and the distribution has
  fatter tails.

3.Platykurtic (Negative Kurtosis):
  A distribution with negative kurtosis has lighter tails and is less peaked
  than a normal distribution.
  It suggests that the data has fewer extreme values, and the distribution is
  more spread out.

In R, the kurtosis of a dataset can be calculated using the `kurtosis()`
function from the `moments` package or the `kurtosis()` function from the
`e1071` package. These functions provide a numerical measure of kurtosis, 
which help in assessing the shape of a distribution.

```{r,tut=TRUE,height=500}
# Using the moments package
# install.packages("moments")

# Load package
library(moments)
data <- c(1, 2, 2, 3, 3, 3, 4, 4, 4, 4)

kurtosis_value <- kurtosis(data)

# Print out value
kurtosis_value
```

In this case we can conclude the data's distribution is leptokurtic.


## SUMMARY FUNCTIONS
Summary functions and describe functions in R are essential tools for
generating descriptive statistics, providing quick insights into the nature of
your data. Here's a breakdown of key functions and considerations:

Base R Summary Functions
`summary()`
- Offers a basic summary of various data types:
- Numeric: Min, max, median, quartiles, mean.
- Factor: Number of levels, counts per level.
- Character: Length of each string.

`str()`
- Displays the internal structure of an object:
- Data type, number of elements, first few values.


Additional Descriptive Statistics Packages:

`psych` package
- `describe()`: Provides a comprehensive overview of descriptive statistics,
    including skewness and kurtosis.


```{r,tut=TRUE,height=500}
# Using the psych package
#install.packages("psych")

# Load package
library(psych)

cars <- datasets::mtcars

# Call describe
describe(cars)
```


`Hmisc` package

describe(): Offers similar descriptive statistics with formatting options.

```{r,tut=TRUE,height=500}
# Using the Hmisc package
#install.packages("misc")

# Load package
library(Hmisc)

# Import cars data
cars <- datasets::mtcars

Hmisc::describe(cars)
```


### Key Considerations for Using Summary Functions

- Data type: The type of summary statistics provided depends on the data type 
    (numeric, factor, character).
- Customization: Some functions allow customization of output and formatting.
- Exploration: Use these functions in conjunction with visualizations 
  (e.g., histograms, boxplots) for a deeper understanding of your data.


Remember:
- Summary and describe functions are invaluable for initial data exploration
    and analysis.
- Choose functions and packages that best suit your data types and analysis
    needs.
- Integrate these functions with visualizations for a comprehensive view of
    your data's characteristics.


## CORRELATION
Correlations are valuable tools in statistics and data analysis, helping us
discover relationships between variables. They quantify the degree of
association between two variables, revealing if they tend to move together
(positive correlation), move in opposite directions (negative correlation),
or have no meaningful relationship (zero correlation).

### Interpreting Correlations
- Correlation coefficient values: Range from -1 (perfect negative correlation)
    to 1 (perfect positive correlation), with 0 indicating no linear
    relationship.
- Strength of association: Consider the absolute value of the coefficient:
    1. Strong: > 0.7
    2. Moderate: 0.3-0.7
    3. Weak: < 0.3
- Direction of association: Positive values indicate variables tend to increase
    together, while negative values show they tend to decrease together.
- Causality vs. correlation: Correlation does not imply causation! Just because
    two variables are associated doesn't mean one causes the other.

Visualizing Correlations:
- Scatter plots: Plot one variable on the x-axis and the other on the y-axis to
    assess the direction and strength of the relationship visually.
- Correlation matrices: Display correlations between all pairs of variables in
    a data frame as a heatmap, showcasing potential patterns and clusters.
    
    
### Common Mistakes
- Assuming causality: Always consider other factors and research theories
    before concluding one variable causes another.
- Ignoring non-linear relationships: Pearson correlation assumes linearity;
    for non-linear relationships, consider Spearman rank correlation or other
    alternatives.
- Focusing only on statistical significance: Even statistically significant
    correlations may not be practically meaningful for your research question.

### Calculating Correlations in R:

- Base R function: cor() calculates the Pearson correlation coefficient,
which measures linear relationships between continuous variables.

```{r,tut=TRUE,height=500}
# Load mtcars
#cars <- read.csv(')

# Preview columns
head(cars)

# Check correlation between mpg and hp
cor(cars$mpg, cars$hp)
```
The correlation value between miles per gallon and horsepower is -0.7761684,
indicating a strong negative correlation between miles per gallon and
horsepower which can be interpreted as, cars with a higher horsepower or more 
powerful engines, tend to have lower fuel efficiency and vice versa.

We can visualize this using a scatterplot like this:

```{r,tut=TRUE,height=500}
# Load mtcars
cars <- datasets::mtcars

# Plot a scatterplot of mpg vs hp
plot(cars$hp, cars$mpg, 
     main = "Scatterplot of hp vs mpg",
     xlab = "Horsepower (hp)",
     ylab = "Miles per Gallon (mpg)",
     pch = 16,
     col = "dodgerblue"
)

```
The scatterplot helps us further see how negatively correlated the two
variables are. We will talk about plots in the next chapter

Alternatives to calculating correlation in R
- `Hmisc::rcorr()` calculates Pearson and Spearman rank correlations for all
    pairs of variables in a data frame.
- `psych::corr.test()` provides correlation coefficients alongside significance
    tests.


## CODE CHALLENGE
1. Using the used car price dataset, find the highest and lowest seats in our
   dataset and assign highest to `max_seats` and `min_seats`.
   
2. Return the variance and standard deviation of mileage in our dataset. Assign
   your results to `mileage_var` and `mileage_sd` respectively.

3. Check for skewness for mileage in our dataset and 
   - if the data is positively skewed, assign "positively skewed" to the
     variable `is_skewed`
   - if the data is positively skewed, assign "negatively skewed" to the
     variable `is_skewed`
   - if it is symmetric, assign "symmetric" to the variable `is_skewed`
   
4. Check for correlation between number of seats and mileage
    - if positively correlated, assign "correlated" to `is_correlated`
    - if negatively correlated, assign "not correlated" to `is_correlated`
    
5. Using the avocado dataset in the datasets folder, check for skewness in
   average avocado prices.
   - if the data is positively skewed, assign "positively skewed" to the
     variable `is_skewed`
   - if the data is positively skewed, assign "negatively skewed" to the
     variable `is_skewed`
   - if it is symmetric, assign "symmetric" to the variable `is_skewed`

6. Find the range of total volume of all avocados and assign to `volume_range`.

7. Report on the standard deviation and variance of the total bags column.
   For standard deviation assign your result to `tot_bags_sd` and for variance,
   assign it to `tot_bags_sd`

8. Find the IQR, median and mean of extra large bags of avocados, assigning
   your results to `xl_bags_iqr`, `xl_bags_median` and `xl_bags_avg`

9. Using the starbucks dataset, return the average calories of beverages that
   were Caffè Latte and assign your result to `latte_avg`

10. For all Classic Espresso Drinks in the beverage category, find the standard
    deviation of total fat they contained and assign to `total_fat_sd`.
