# GRAPHS

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

<script src="files/js/dcl.js"></script>
```{r ,include=FALSE}
tutorial::go_interactive(greedy=TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```

Drawing graphs is a fundamental and powerful tool in data analysis and
visualization, enabling the representation of complex relationships and
patterns within datasets. Graphs provide a visual means of conveying
information, making it easier for individuals to interpret and understand
trends, comparisons, and distributions. Whether creating scatter plots, bar
charts, line graphs, or more sophisticated visualizations, the process of
drawing graphs allows analysts to communicate insights effectively and
facilitates the exploration of data structures. 

In various fields, from scientific research to business analytics, the art of
drawing graphs plays a pivotal role
in transforming raw data into actionable insights, supporting evidence-based
decision-making and effective communication of findings to diverse audiences.

## GRAPHING/VISUALIZATION WITH DIFFERENT PACKAGES
There are countless visualization packages in R however, some of the most
popular ones include:

1. Base R Graphics:
  - Built-in functions like `plot()`, `hist()`, `boxplot()`, and `barplot()`
    offer basic visualizations.
  - Simple to use but limited customization options.
  - Suitable for quick exploratory plots or basic reports.
2. ggplot2:
  - The king of R visualization!
  - A layered grammar-based approach, providing incredible flexibility and
    control over every aspect of the plot.
  - Endless possibilities for creating publication-quality graphics with
    diverse themes and data transformations.
  - Requires some learning curve but offers immense rewards.
3. lattice:
  - Offers a powerful framework for creating multi-panel lattice plots and
    trellis plots.
  - Useful for comparing several groups or variables within a single graphic.
  - More complex than base R graphics but less demanding than ggplot2 for
    certain types of visualizations.
4. plotly:
  - Creates interactive web-based visualizations.
  - Allows users to explore and manipulate the data through zoom, hover, and
    other interactive features.
  - Ideal for sharing findings or presenting data in a dynamic way.
5. cowplot:
  - Facilitates easy combination of multiple plots into a single layout.
  - Useful for creating composite images with various types of graphs
    side-by-side or overlaid.
  - Simplifies presenting a multifaceted analysis in a single visual.

## Data Visualization Examples
For the purpose of this course, we will discuss examples using base R and
`ggplot2`.

### Scatterplot
A scatterplot is a graphical representation that reveals the relationship
between two continuous variables. By plotting points on a two-dimensional
plane, each point corresponds to a pair of values. This visualization is
particularly useful for identifying patterns, trends, and potential
correlations within the data. Scatterplots are effective when exploring
numerical data

#### Base R
In base R, we can build scatterplots like this
```{r,tut=TRUE,height=500}
# Load mtcars
cars <- datasets::mtcars

# With Base R
plot(x = cars$disp,
     y = cars$hp,
     col = "steelblue")
```

#### ggplot2
In `ggplot2`, the syntax for building scatterplots is different, with more
options for customization. ggplot2 has different arguments as the second layer
for each plot type. For a scatterplot in `ggplot2` the second layer to specify
is `geom_point()`. You can create a scatterplot like this:

```{r,tut=TRUE,height=500}
# With ggplot2
library(ggplot2)

ggplot(data = cars, aes(
  x = disp,
  y = hp
)) +
  geom_point(color="steelblue") +
  theme_minimal()
```


### Histogram
Histograms are used to show the distribution of a variable. Typically they are
used for numerical data, histograms present data in intervals or bins along the
x-axis, with the frequency or density of observations depicted by the height of
the bars on the y-axis. This visualization aids in understanding the central
tendency, spread, and shape of the data distribution, providing insights into
the data's overall pattern.

#### Base R

```{r,tut=TRUE,height=500}
# Load mtcars
cars <- datasets::mtcars

# With Base R
hist(cars$hp,
     col="darkgreen")
```


#### With ggplot2
The argument to supply here after creating the first layer is
`geom_histogram()`
```{r}
# Load mtcars
cars <- datasets::mtcars

# With ggplot2
library(ggplot2)

ggplot(data = cars, aes(
  x = hp
)) +
  geom_histogram(color="black", fill="darkgreen") +
  labs(
    x = "Horsepower(hp)",
    title = "Histogram of Horsepower"
  ) +
  theme_minimal()
```


### Boxplot
Boxplots provide a concise summary of the distribution and key statistics of a
dataset. Ideal for numerical data, boxplots display the median, quartiles, and
potential outliers in a compact visual format. They are especially valuable for
comparing distributions between different groups and identifying variations in
central tendency and spread. Boxplots are particularly effective when dealing
with large datasets or when a quick overview of data is needed.


#### Base R

```{r,tut=TRUE,height=500}
# Load mtcars
cars <- datasets::mtcars

# With Base R
boxplot(mpg ~ carb, 
        data = mtcars, 
        col = "lightcoral", 
        main = "Boxplot of mpg by Carburetor Type", 
        xlab = "Carburetor Type", ylab = "Miles per Gallon")
```

#### ggplot2

```{r}
# Load mtcars
cars <- datasets::mtcars

# With ggplot2
library(ggplot2)

ggplot(data = cars, aes(
  x = factor(carb),
  y = mpg,
  fill = factor(carb)
)) +
  geom_boxplot(color="black", fill = "lightcoral") +
  labs(
    x = "Carburetor Type",
    y = "Miles per Gallon",
    title = "Boxplot of mpg by Carburetor Type"
  ) +
  theme_minimal()
```


### Barplots
Barplots present categorical data with rectangular bars, where the length of
each bar corresponds to the frequency or proportion of the category it
represents. Barplots are effective for comparing the relative sizes of
different categories within a dataset.

#### Base R
```{r,tut=TRUE,height=500}
# Load mtcars
cars <- datasets::mtcars

# Create a table of the frequency of each gear type
gear_table <- table(cars$gear)

# With Base R
barplot(gear_table,
        col = "gold", 
        main = "Barplot of Gears", 
        xlab = "Gear type", ylab = "Frequency")
```

#### ggplot2
In ggplot2, geom_bar and geom_col functions are two commonly used geometries in
`ggplot2` for creating barplots. Let's explore the differences and how to build
a barplot using the cars dataset.

##### geom_bar vs. geom_col
`geom_bar`
  - geom_bar is used when you want the height of the bar to represent the count
  of cases in each group.
  - By default, geom_bar uses stat="count", and it counts the number of cases
  in each group (in this case, the frequency of each gear type).
  
We can build a barplot with `geom_bar()` like this:

```{r}
# Load mtcars
cars <- datasets::mtcars

# With ggplot2
library(ggplot2)

# Using geom_bar
ggplot(cars, aes(x = factor(gear))) +
  geom_bar(color = "black", fill = "gold") +
  labs(
    title = "Barplot of Gears",
    x = "Gear type",
    y = "Frequency"
  ) +
  theme_minimal()
```


`geom_col`
  - geom_col is used when you want to specify the height of the bars directly,
  instead of relying on the default count.
  - You should use geom_col when you have pre-computed heights or when the data
  is already summarized.


We can build a barplot with ggplot2 in R with `geom_col()` like this.

```{r}
# Load mtcars
cars <- datasets::mtcars

# With ggplot2
library(ggplot2)

# Using geom_col without precomputing
ggplot(cars, aes(x = factor(gear))) +
  geom_col(fill = "gold", color = "black") +
  labs(
    title = "Barplot of Gears",
    x = "Gear type",
    y = "Frequency"
  ) +
  theme_minimal()
```


### Pie chart
Pie charts are used to display the proportional contribution of each
category to a whole. This visualization is particularly effective when
presenting parts of a whole and providing a quick overview of the distribution
of categorical data. Each slice of the pie represents a category, with its size proportional to the share it contributes to the whole. While widely used, it's
essential to be cautious with pie charts, especially when dealing with a large
number of categories, as they can become difficult to interpret accurately.

#### Base R

```{r,tut=TRUE,height=500}
# Load mtcars
cars <- datasets::mtcars

# Create a table of the frequency of each cylinder type
cyl_table <- table(cars$cyl)

# With Base R
pie(cyl_table, 
    main = "Pie Chart of Car Models by Cylinder Type",
    col = rainbow(length(cyl_table)),
    labels = paste(names(cyl_table), ": ", cyl_table),
    cex = 0.8)
```

#### ggplot2
```{r}
# Load mtcars
cars <- datasets::mtcars

# With ggplot2
library(ggplot2)

# Convert the table to a data frame
cyl_df <- as.data.frame(cyl_table)
colnames(cyl_df) <- c("Cylinders", "Frequency")


# Create a bar plot with polar coordinates
ggplot(cyl_df, aes(x = "", y = Frequency, fill = factor(Cylinders))) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  theme_void() +
  labs(
    title = "Pie Chart of Car Models by Cylinder Type") +
  scale_fill_discrete(name = "Number of Cylinders")
```


There is a slight syntax difference for bar charts in ggplot2. Read more about
it [here](https://mpn.metworx.com/packages/ggplot2/3.3.0/reference/geom_bar.html#:~:text=There%20are%20two%20types%20of,data%2C%20use%20geom_col()%20instead.)


### Dotplots
Dotplots are a straightforward method to visualize the distribution of a
dataset, particularly useful for smaller datasets. Individual dots are placed
along a number line, with each dot representing an observation. Dotplots
provide a clear representation of the data points and help identify patterns
and clusters. While not as common as some other types of visualizations,
dotplots can be valuable when a simple, uncluttered display of data is desired,
facilitating a direct and intuitive interpretation.

```{r,tut=TRUE,height=500}
# Load mtcars
cars <- datasets::mtcars

# Count the frequency of each cylinder type
cyl_counts <- table(cars$cyl)

# With Base R
plot(cyl_counts, type = "h", 
     lwd = 10, col = "lightblue", 
     xaxt = "n", main = "Dot Plot of Car Models by Cylinder Type", 
     xlab = "Number of Cylinders", ylab = "Count")
axis(1, at = seq_along(cyl_counts), labels = names(cyl_counts))

```


```{r}
# Load mtcars
cars <- datasets::mtcars

# With ggplot2
library(ggplot2)

# Create a dot plot
ggplot(cars, aes(x = factor(cyl))) +
  geom_dotplot(fill = "lightblue") +
  labs(
    title = "Dot Plot of Car Models by Cylinder Type",
    x = "Number of Cylinders",
    y = "Count"
  )
```


## WHICH VISUALIZATION PACKAGE TO USE?
- Relevance: Base R for quick checks, ggplot2 for extensive customization, 
    lattice for multi-panel plots, plotly for interactivity,
    cowplot for combined visuals.
- Audience: Base R graphics might suffice for technical reports,
    while ggplot2 or plotly cater to broader audiences who benefit from visual
    clarity and interactivity.
- Learning curve: Start with simpler packages like base R if new to R,
    then graduate to ggplot2 or others for advanced visualizations.

## CODE CHALLENGE
You are to use base R and ggplot2 for each question below. For the next 5
questions, using the CO2 dataset from the `datasets` package:
1. Visualize the distribution of concentration levels and assign the base R
   visualization to `base_viz1` and ggplot one to `gg_plot1`

2. Visualize the distribution of uptake by type using a boxplot and assign the
   base R visualization to `base_boxplot` and ggplot one to `gg_boxplot`

3. For each plant, visualize the concentration levels using a barplot. Assign
   the base R viz to `base_bar` and ggplot to `gg_bar` respectively

4. Create a scatterplot to show the relationship between concentration and
   uptake levels and assign the base R and ggplot scatterplots to 
   `base_scatterplot` and `gg_scatterplot` respectively
   
5. Visualize the distribution of sepal length in base R and petal width with
   ggplot2 using the iris dataset and assign the base R visualization to
   `base_viz2` and ggplot one to `gg_plot2`
   
6. Visualize the distribution of petal lengths by species using a boxplot using
   the iris dataset and assign the base R visualization to `base_iris_boxplot`
   and ggplot one to `gg_iris_boxplot`
   
7. Visualize the distribution of car weights by the number of cylinders using
   the mtcars dataset and assign the base R visualization to
   `base_viz3` and ggplot one to `gg_plot3`
   
8. Create a scatterplot to show the relationship between car weight and miles
   per gallon and assign the base R and ggplot scatterplots to 
   `base_car_scatterplot` and `gg_car_scatterplot` respectively
   
9. Using the airquality dataset, visualize the wind speed for each day using a
   barplot. Assign the base R viz to `base_wind_bar` and ggplot to
   `gg_wind_bar` respectively
   
10. Visualize the distribution of sepal widths by species using the iris
    dataset and assign the base R visualization to `base_iris_viz` and ggplot
    one to `gg_iris_plot`

In your pull request, add the plot images and the script used to create these
plots. Use the ggsave function to save your plots. This [material](https://ggplot2.tidyverse.org/reference/ggsave.html) will help you.
