# FUNCTIONS
<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

<script src="files/js/dcl.js"></script>
```{r ,include=FALSE}
tutorial::go_interactive(greedy=TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```

In R, functions are reusable blocks of code that perform specific tasks. They
are essential for organizing, streamlining and making your code more readable
and maintainable. Functions in R typically take in input parameters, perform 
operations based on those inputs and return a result.


## CREATING A FUNCTION
Let us see the basic structure of a function.

```{r,tut=TRUE,height=500}
# Basic structure of a function
function_name <- function(parameter1, parameter2, ...) {
  # Function body - code to perform the task
  # ...
  
  # Return statement (optional) - specifies the value to be returned
  # return(some_value)
}

# Call a function
result <- function_name(parameter)

```

### Overview of a basic function structure
Here is an overview of the basic structure of a function:
- function_name: This is the name you give to your function.
- parameters: These are variables that represent the input values passed to the
  function. You can have zero or more parameters.
- Function body: This is the block of code inside the function that performs
  the desired operations.
- Return statement: (optional) This statement specifies the value that the
  function will return. If no return statement is provided, the function
  returns the result of the last evaluated expression.
  

To call a function, we will just need to type the function name, that is, the
variable the function was assigned to and then add a bracket and add a
parameter or parameters where applicable.

## Examples of functions

1. Creating a function that calculates the area of a rectangle.

Whenever creating a function, it is important to break down what you want
your function to do.

We want to create a function that can calculate the area of a rectangle. From 
basic mathematics, the area of a rectangle can be calculated from this:
`Area = length * width`

Let us create a function to calculate the area of a rectangle programmatically
in R.

```{r,tut=TRUE,height=500}
# Function to calculate area of a rectangle
calculate_area <- function(length, width) {
  # Multiply length by width and assign to variable called area
  area <- length * width
}

# Call function
rectangle <- calculate_area(5,15)

# Print rectangle
rectangle
```
This works. But when creating complex functions, due to expected behavior, it
is always recommended to use return explicitly to return a value. Given this,
let us properly define our calculate_area function.

```{r,tut=TRUE,height=500}
# Function to calculate area of a rectangle
calculate_area <- function(length, width) {
  # Multiply length by width and assign to variable called area
  area <- length * width
  
  # Explicitly return area
  return(area)
}

# Call function
rectangle <- calculate_area(5,15)

# Print rectangle
rectangle
```
Now we have defined our function properly to return a value.


2. Creating a basic function to add two numbers
We will create a function that takes two numbers and returns the sum.

```{r,tut=TRUE,height=500}
# Function to add two numbers
add_numbers <- function(x, y) {
  result <- x + y
  
  # Return result
  return(result)
}

# Call function and assign result to a variable
sum_result <- add_numbers(3, 5)

# Print result
sum_result
```


## FUNCTIONS WITH DEFAULT PARAMETERS
Sometimes, some functions can take on default parameters and can still run
without throwing up an error.

Let us revisit our calculate area function again and let us call it without
adding any parameters.

```{r,tut=TRUE,height=500}
# Function to calculate area of a rectangle
#calculate_area <- function(length, width) {
#  # Multiply length by width and assign to variable called area
#  area <- length * width
#  
  # Explicitly return area
#  return(area)
#}

# Call function
#calculate_area()
```
Error! We get the error: "Error in calculate_area() : argument "length" is 
missing, with no default"

This is because we did not add default parameters to our function. This is a
form of error-handling. For now let us add default  parameters to our function
which will ensure our function runs even when no values are supplied.

```{r,tut=TRUE,height=500}
# Function with default values for parameters
calculate_area_with_defaults <- function(length = 1, width = 1) {
  area <- length * width
  return(area)
}

# Call function without specifying arguments
calculate_area_with_defaults()
```
Our function now returns 1 and does not return an error! Defining functions 
with a default parameter or parameters ensure our function do not return errors
when run without parameters. 

There are other aspects of functions you should pay attention to when creating
functions. You need to ensure you implement some error-handling features to
handle some unexpected behaviors in your function. For example in our 
`calculate_area()` function, we can add an error-handling feature that ensures
the parameters cannot be zero since multiplying any number by zero returns 
zero.

Let us implement an error handling feature that makes sure we do not have any
zero or negative parameters.

```{r,tut=TRUE,height=500}
# Error handling example
#calculate_area_safe <- function(length, width) {
#  if (length <= 0 || width <= 0) {
#    stop("Length and width must be positive values.")
#  }
#  area <- length * width
#  return(area)
#}
#
# Call function with error
#error_rectangle1 <- calculate_area_safe(-2, 8)
#error_rectangle2 <- calculate_area_safe(5, 0)

# Print result
#error_rectangle
#error_rectangle2
```

Good. Supplying a negative number and also 0 returns errors. We can make sure
we handle unexpected behavior with our function.

## CODE CHALLENGE
1. Create a function called `check_even_odd` that takes in a given positive
   integer and checks if it is an even or odd number.
   
2. Create a function named `is_prime` that checks if a given positive integer
   is a prime number. The function should return a logical value (TRUE or FALSE)
   
3. Given a triangle with a base and a height, create a function called
   `calculate_triangle_area` that calculates the area of said triangle 
   
4. Create a function called find_maximum that takes in two numbers as arguments
   and returns the larger of the two.
   
5. Create a function named `calculate_average` that takes a numeric vector as
   an argument and returns the average (mean) of the values.

6. Develop a function named `is_palindrome` that checks if a given string is a
   palindrome. A palindrome is a word, phrase, or sequence of characters that
   reads the same forward and backward, ignoring spaces, punctuation, and
   capitalization. Use `tolower()` to convert any input to lower cases inside
   your function code before you do anything else.

7. Create a function named `calculate_factorial` that takes a positive integer
   as an argument and returns the factorial of that number. The factorial of a
   non-negative integer n is the product of all positive integers less than or
   equal to n. Read more about factorial [here](https://en.wikipedia.org/wiki/Factorial)

8. Create a function called `is_positive` that takes a numeric value as an
   argument and returns TRUE if the number is positive, and FALSE otherwise.
   
9. Implement a function named `is_leap_year` that takes a year (as a positive
   integer) as an argument and returns TRUE if it's a leap year and FALSE
   otherwise. A leap year is divisible by 4, but not divisible by 100 unless it
   is also divisible by 400.

10. Create a function named `reverse_vector` that takes a vector as an argument
   and returns a new vector with the elements reversed. The function should not
   modify the original vector, only reverse the order of elements.
