# GROUP MANIPULATION

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

<script src="files/js/dcl.js"></script>
```{r ,include=FALSE}
tutorial::go_interactive(greedy=TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```

Group manipulation in R refers to operations or transformations applied to data
grouped by one or more variables.

## APPLY FAMILY
The apply family in R comprises a set of powerful and versatile functions
designed for operations on matrices, arrays, lists, and vectors. These
functions, including `apply()`, `lapply()`, `sapply()`, `vapply()`, and
`mapply()`, allow us to apply a specified function to the rows or columns of a
matrix, elements of a list or vector, or multiple vectors simultaneously.

These functions are commonly used in tasks such as calculating summary
statistics, performing transformations, or building models within specific
groups of data.

### The Apply Functions
Understanding the Family:
- apply(): The most general function, allowing you to specify the margins
  (rows, columns, etc.) over which to apply a function.
- lapply(): Applies a function to each element of a list, returning a new
  list with the transformed elements.
- sapply: Similar to lapply, but simplifies the return value to a vector or
  matrix.
- vapply(): Like sapply, but applies a function by variable instead of element,
  useful for data frames.
- tapply(): Specifically designed for group manipulation, applies a function to
  each group defined by a factor variable.

### apply()
The `apply()` function is a powerful tool for applying a function repeatedly to
elements of an array or matrix, often eliminating the need for explicit loops.

It offers flexibility and control over how the function is applied, making it
a extremely efficient in data manipulation.

SYNTAX: `apply(X, MARGIN, FUN, ...)`

- X: The array or matrix on which the function will be applied.
- MARGIN: Specifies the margin or dimension along which the function is applied
  (1 for rows, 2 for columns).
- FUN: The function to be applied.
- ...: Additional arguments to be passed to the function.

```{r,tut=TRUE,height=500}
# Sum the elements of each column in a matrix
matrix_data <- matrix(1:6, nrow = 2)

# Apply sum function to matrix_data
column_sums <- apply(matrix_data, 2, sum)

column_sums
```

Let us take another look at an example where we use `apply()` to multiply the
elements of each row of a matrix

```{r,tut=TRUE,height=500}
# Define a matrix
matrix_data <- matrix(1:6, nrow = 2)

# Apply multiplication function to matrix_data along rows
row_products <- apply(matrix_data, 1, prod)

row_products
```

Can you try to use `apply()` to calculate the mean of each column in a matrix?
The matrix is already predefined below

```{r,ex="apply",type="sample-code"}
# Create a matrix


# Calculate the mean of each column

```

```{r,ex="apply",type="solution"}
# Create a matrix
matrix_data <- matrix(1:6, nrow = 2)

# Calculate the mean of each column
col_means <- apply(matrix_data, 2, mean)
col_means
```


### lapply()
lapply stands for "list apply" and is used to apply a function to each element
of a list. It returns a list where each element is the result of applying the
specified function to the corresponding element of the input list.

SYNTAX: `lapply(X, FUN, ...)`

```{r,tut=TRUE,height=500}
# Create a named list
num_list <- list(a = 1, b = 2, c = 3)

# Square each element of a list
squared_list <- lapply(num_list, function(x) x^2)

# Print result
squared_list
```

We can also use `lapply()` on character type lists. Let us consider an example
below where we convert each list element from lower to uppercase.

```{r,tut=TRUE,height=500}
# Create a character list
char_list <- list(name1 = "john", name2 = "jane", name3 = "bob")

# Convert each character element to uppercase
uppercase_list <- lapply(char_list, function(x) toupper(x))

uppercase_list

```

You might come across various use cases where you will find `lapply()` very
handy.


### sapply()
Similar to lapply, sapply stands for "simplified apply." It applies a function
to each element of a list but attempts to simplify the result into a vector or
matrix if possible.


SYNTAX: `sapply(X, FUN, ...)`

```{r,tut=TRUE,height=500}
# Square each element of a list and simplify the result
num_list <- list(a = 1, b = 2, c = 3)
squared_vector <- sapply(num_list, function(x) x^2)

# Print result
squared_vector
```

We could also decide to use `sapply()` to double the elements in a numeric list
for example. We can do that like this

```{r,tut=TRUE,height=500}
# Create a numeric list
num_list <- list(a = 4, b = 5, c = 6)

# Double each element of the list and simplify the result
doubled <- sapply(num_list, function(x) x * 2)

# Print result
doubled

```

As you can see, `sapply()` simplifies our output into a vector again. In case
you want to remember what it does, it is used to perform operations on lists
and tries to return a vector if it is possible.

`sapply()` can also be used for lists containing wholly string data types.

Let us say we had a list of names. We can use `sapply()` to concatenate 
"Hello, " before each name. How can we do that? Easy. We can use `sapply()` to
apply base R's `paste()` function to do the job for us like this:

```{r,tut=TRUE,height=500}
# Create a character list
char_list <- list(name1 = "Fatima", name2 = "Bob", name3 = "Steve")

# Concatenate "Hello, " to each character element and simplify the result
greeting <- sapply(char_list, function(x) paste("Hello, ", x))

greeting

```


### vapply()
vapply is similar to sapply but allows you to specify the type and shape of the
result, providing a safer and more predictable output.
SYNTAX: `vapply(X, FUN, FUN.VALUE, ...)`

```{r,tut=TRUE,height=500}
# Square each element of a vector and specify the result type
num_vector <- c(1, 2, 3)
squared_vector <- vapply(num_vector, function(x) x^2, numeric(1))

# Print result
squared_vector
```

### tapply()
`tapply()` is specifically designed for group-wise operations. It applies a
function to subsets of a vector defined by a factor variable

SYNTAX: `tapply(X, INDEX, FUN, ...)`

```{r,tut=TRUE,height=500}
# Calculate the mean of a numeric vector for each group defined by a factor
values <- c(10, 23, 13, 84, 50)
groups <- factor(c("A", "B", "A", "B", "A"))

# Calculate the mean
mean_by_group <- tapply(values, groups, mean)

# Print result
mean_by_group
```


## aggregate()
In R, the aggregate function is another powerful tool for manipulating data, 
specifically focused on calculating summary statistics for subsets of a data 
frame based on grouping variables. Think of it as a specialized tool for
calculating group means, medians, counts, etc. The aggregate function can help
replace cumbersome loops and manual calculations for grouped statistics, it is
very flexible and allows a user more control over specific variables or
grouping factors.

SYNTAX: `aggregate(x, by, FUN, ...)`

### Arguments
Arguments:
- x: The variable(s) for which you want to calculate the summary statistics.
- by: A list of columns containing the grouping variables.
- FUN: The function to apply to each group. Built-in options like mean, median,
sum, etc., or your own custom function.
- ...: Additional arguments passed to the FUN function.


Let us see the `aggregate()` function in action.

```{r,tut=TRUE,height=500}
# Load chickwts data
chickweight <- datasets::chickwts

summary_stats <- aggregate(weight ~ feed, data = chickweight, mean)

# Print out result
summary_stats
```

`aggregate()` is flexible and can be used to calculate multiple statistics by
providing a vector of functions for the `FUN` argument.

`aggregate()` might be a complex for new R users to fully grasp and it is
usually recommended for new users to use `dplyr`'s `group_by` and `summarize()`
functions due to its more readable syntax.


## plyr
plyr is an R package that provides a set of tools for splitting, applying, and
combining data in a systematic way. 

This pattern is common in data analysis: you break down a complex problem into
smaller pieces, manipulate each piece, and then combine the results back
together.

The plyr package introduces a family of functions that follow a common pattern
for data manipulation. The key functions in the plyr package include:

1. ddply: This function is used for splitting a data frame, applying a function
to each subset, and then combining the results into a data frame. It's
particularly useful for tasks involving grouping and summarizing data.

```{r,tut=TRUE,height=500}
library(plyr)

# Syntax
#ddply(data_frame,
#      .(grouping_variable), 
#      summarise, 
#      mean_value = mean(numeric_variable))

# Load the mtcars dataset
data(mtcars)

# Use ddply to calculate the mean and sum of mpg by the number of cylinders
ddply(mtcars, .(cyl), summarise, mean_mpg = mean(mpg), sum_mpg = sum(mpg))

```

2. ldply(): It is used for applying a function to each element of a list and
then combining the results into a data frame. This is handy when working with
lists of data.

```{r,tut=TRUE,height=500}
# Load plyr
library(plyr)

# Syntax of ldply()
#ldply(list_of_data_frames, 
#      function(x) summarise(x, mean_value = mean(numeric_variable)))


# Create a list of data frames (you can replace this with your own list)
list_of_data_frames <- list(
  data.frame(id = 1:3, value = c(10, 15, 20)),
  data.frame(id = 4:6, value = c(25, 30, 35))
)

# Apply ldply
ldply(list_of_data_frames, function(x) summarise(x, mean_value = mean(value)))
```

3. adply(): Similar to ddply, but it returns an array or matrix as the result.

```{r,tut=TRUE,height=500}
# Syntax of adply()
#adply(data_frame, 
#      .(grouping_variable), 
#      function(x) c(mean_value = mean(x$numeric_variable)))
```


### Other Key plyr functions

- plyr: The core function for "split-apply-combine" operations.
- mutate: Add or modify columns within each data subset.
- rename: Rename columns after applying functions.
- arrange: Reorder rows based on specified columns.
- summarize: Create a new data frame with summary statistics for each group.
- join: Combine data subsets based on shared variables.



It would be interesting to note that, dplyr, is now preferred over plyr due to
its more readable syntax and modern approaches to statistical computing.


## RESHAPING
In R, reshaping refers to the process of restructuring or transforming the
layout of your data, often changing it from one form to another. This is
particularly useful when working with data in different formats, and it's a
common step in data preparation and analysis.

Reshaping is typically done using functions that rearrange the rows and
columns of your data to make it more suitable for your analysis or
visualization needs.


There are two primary types of reshaping in R and they are:
- wide to long reshaping
- long to wide reshaping

### Wide to Long
Wide to long reshaping transforms a dataset from a wide-format dataset, where
each variable has its own column, into a long-format dataset, where all the
variable values are stacked in a single column with an additional column
indicating the variable name.

We can use two functions to achieve this:
- `pivot_longer()` from the `tidyr` package
- `gather()` from the `dplyr` package.

```{r,tut=TRUE,height=500}
# Using tidyr::pivot_longer
library(tidyr)
# Syntax
#long_data <- pivot_longer(wide_data, cols = starts_with("variable_prefix"),
#names_to = "variable_name", values_to = "value")

# Loard iris dataset
iris <- datasets::iris

# Wide to Long Reshaping using tidyr::pivot_longer
long_iris <- pivot_longer(iris, cols = -Species, names_to = "Variable",
                          values_to = "Value")

# Preview transformed dataframe
head(long_iris)

```


```{r,tut=TRUE,height=500}
# Using dplyr::gather
library(dplyr)


#long_data <- gather(wide_data,
#                    key = "variable_name", 
#                    value = "value", 
#                    starts_with("variable_prefix"))

# Create a simple wide-format dataset
wide_data <- data.frame(
  ID = 1:5,
  var_A = c(10, 20, 30, 40, 50),
  var_B = c(15, 25, 35, 45, 55),
  var_C = c(22, 32, 42, 52, 62)
)

# View the original wide-format dataset
wide_data

# Wide to Long Reshaping using dplyr::gather
long_data <- gather(wide_data, key = "variable_name", value = "value", -ID)

# View the reshaped long-format dataset
long_data

```

### Long to Wide
Transforms a long-format dataset into a wide-format dataset, where each unique
value in a specified column becomes a new column.

We can use:
- `pivot_wider` from `tidyr`
- `spread` from `dplyr`

```{r,tut=TRUE,height=500}
# Using tidyr::pivot_wider
library(tidyr)

# Pivot wider 
# wide_data <- pivot_wider(long_data, names_from = "variable_name",
#values_from = "value")

# Load mtcars
mtcars <- datasets::mtcars

# Long to Wide Reshaping using tidyr::pivot_wider
wide_data_tidyr <- pivot_wider(mtcars, names_from = "gear",
                               values_from = "mpg")

# View the reshaped wide-format dataset
wide_data_tidyr
```

For the `spread()` function from the `dplyr` package.
```{r,tut=TRUE,height=500}
# Using dplyr::spread
library(dplyr)

# Pivot from long to wide
#wide_data <- spread(long_data, key = "variable_name", value = "value")

# Load mtcars
mtcars <- datasets::mtcars

# Long to Wide Reshaping using dplyr::spread
wide_data_dplyr <- spread(mtcars, key = "gear", value = "mpg")

# View the reshaped wide-format dataset
wide_data_dplyr
```


## CODE CHALLENGE
1. Create a list named fruits containing the elements: apple, mango, peach,
   orange, watermelon, avocado, and pawpaw. Use `lapply()` to apply an
   anonymous function that counts the number of characters in each element. 
   Assign the resulting list to a variable named `fruits_char_counts`
   
2. Using the chickwts dataset, calculate the average weight for each feed
   category using the `tapply()` function. Assign the resulting averages to a
   variable named `feed_avg_weights`
   
3. Reshape the airquality dataset from wide to long format. Assign the
   resulting long format dataset to a variable named `airquality_long`

4. Using the `sleep` dataset(use `data(sleep)` to load it), transform it
   from long to wide format and assign the resulting wide format dataset to 
   a variable named `sleep_wide`
   
5. Use the aggregate() function to group the mtcars dataset by the number of
   cylinders and calculate the average miles per gallon for each cylinder
   category. Assign the resulting aggregated dataset to a variable named
   `avg_mpg_by_cyl`
6. Use the aggregate() function to calculate the average sepal length
   for each species (Species). Assign the resulting aggregated dataset to a
   variable named `avg_sepal_length_by_species`
   
7. Create a list named `numeric_lists` containing numeric vectors, vector 1
   contains values from 25 to 32, the second vector contains values from 41 to
   48 and the third vector contains values from 7 to 14. Use `lapply()` to
   apply an anonymous function that squares each element in each numeric
   vector. Assign the resulting list to a variable named `squared_values`

8. Create a list named `age_lists` containing three numeric vectors:
   c(17, 21, 3, 19, 50), c(14, 50, 77, 16, 87), c(37, 8, 90, 43, 60). Use
   `sapply()` to apply an anonymous function that calculates the mean of each
   ages vector. For simplification, name the vectors, v1, v2 and v3
   respectively. Assign the resulting vector of means to a variable
   named `means`

9. Given a matrix `mat`, calculate the mean of non-missing values in each
   column using `apply()`. Assign the resulting vector of column-wise means to a
   variable named `column_means`

10. Given a list named `scores` calculated out of 10 containing numeric
    vectors, use `lapply()` to calculate the cumulative sum for each vector.
    Assign the resulting list of cumulative sums to a variable named
    `other_cumulative_sums`
