# GENERAL STATISTICS

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

<script src="files/js/dcl.js"></script>
```{r ,include=FALSE}
tutorial::go_interactive(greedy=TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```

## TABULATING FACTORS AND CREATING CONTINGENCY TABLES


### Tabulating Factors
Tabulating factors involves creating frequency tables to summarize the
distribution of categorical variables. It provides a quick overview of the
count or percentage of observations in each category. R provides various
functions for tabulating factors, and two commonly used ones are `table()` and
`xtabs()` and `prop.table()`

### Tabulation using table()
The `table()` function in R is a simple and versatile way to create contingency
tables, which displays the frequency counts of combinations of factor levels.
It takes one or more factors as arguments and returns a table of counts.

```{r,tut=TRUE,height=500}
# Create a simple categorical vector
gender <- c("Female", "Female", "Male", "Female", "Male", "Female", "Male",
            "Male", "Female", "Female", "Female", "Male", "Female", "Female",
            "Male")

# Tabulate the gender variable
gender_table <- table(gender)

# Display the table
gender_table
```

You can also specify two factors like this:

```{r,tut=TRUE,height=500}
# Load the mtcars dataset
data(mtcars)

# Frequency table for number of cylinders and gear type
table(mtcars$cyl, mtcars$gear)
```


### Tabulation using xtabs()
The `xtabs()` function is used to create contingency tables similar to
`table()` but it also allows for the inclusion of a formula argument. This
formula approach provides more flexibility in specifying the relationships
between factors.

```{r,tut=TRUE,height=500}
# Create a data frame with two categorical variables
data <- data.frame(
  gender = c("Male", "Female", "Male", "Male", "Female", "Female", "Male",
             "Female", "Male", "Male", "Female", "Female"),
  category = c("Democrat", "Republican", "Republican", "Republican",
               "Democrat", "Democrat", "Republican", "Democrat", "Republican",
               "Republican", "Democrat", "Democrat")
  )

# Tabulate the gender and category variables
cross_table <- xtabs(~ gender + category, data)

# Display the cross-tabulation
cross_table

```

```{r,tut=TRUE,height=500}
# Load the mtcars dataset
data(mtcars)

# Contingency table
xtabs(~ cyl + vs, data = mtcars)
```


Try in the code chunk below to create a contingency table with the variables
`cyl` and `am`
```{r,tut=TRUE,height=500}
# Load the mtcars dataset
data(mtcars)

# Contingency table
xtabs(~ cyl + vs, data = mtcars)
```


### Tabulation with `prop.table()`
The `prop.table()` function is used to compute proportions or percentages from
contingency tables. It takes a contingency table as its main argument and
allows you to specify whether you want proportions across rows, columns, or 
the entire table.

```{r,tut=TRUE,height=500}
# Load the mtcars dataset
data(mtcars)


# Example using prop.table()
prop_table_row <- prop.table(table_cyl_vs_gear, margin = 1)  # Proportions across rows
prop_table_col <- prop.table(table_cyl_vs_gear, margin = 2)  # Proportions across columns
prop_table_total <- prop.table(table_cyl_vs_gear)    
```

### WHEN TO USE WHICH TABULATION FUNCTION
1. Formula vs. Argument
   `table()` takes factors as direct arguments, while `xtabs()` allows the use
   of formulas for more complex relationships between factors.

2. Output Format:
  Both `table()` and `xtabs()` produce contingency tables, but the structure of
  the output can differ slightly.

3. Proportions
  `prop.table()` is specifically designed for computing proportions from
  contingency tables. It allows you to calculate proportions across rows,
  columns, or the entire table.

4. Flexibility:
  While `table()` is straightforward and commonly used for basic tabulation,
  `xtabs()` provides more flexibility when dealing with complex relationships
  between factors.

##  z-Scores
Z-scores, also known as standard scores, measure how many standard deviations a
specific point is away from the mean of a population or sample. They are
calculated by subtracting the mean from the individual value and then dividing
by the standard deviation.


### Applications
- Identifying outliers: Points with z-scores greater than 3 (or less than -3)
  are considered outliers and may warrant further investigation.
- Comparing data points from different distributions: Z-scores allow for
  standardized comparisons even when   data sets have different means and
  standard deviations.
- Normalizing data for statistical tests: Many statistical tests assume
  normality, and z-scores can transform non-normal data into a more normal
  distribution.

### Calculation
z = (x - μ) / σ

where:
- x is the individual value
- μ is the population or sample mean
- σ is the population or sample standard deviation

We can replicate this in R like this:

```{r,tut=TRUE,height=500}
# Create a vector of values
values <- c(10, 15, 20, 25, 30)

# Calculate zscore
z_scores <- (values - mean(values)) / sd(values)


# Print the results
print(z_scores)

```


```{r,tut=TRUE,height=500}
hist(z_scores)  # Histogram to visualize distribution
qqnorm(z_scores)  # Normal quantile-quantile plot to assess normality
```


## TESTING THE MEAN OF A SAMPLE (t-Test) AND ITS CONFIDENCE INTERVAL
The t-test is a statistical test used to determine if the mean of a sample is
significantly different from a hypothesized population mean. It is used when
the population standard deviation is unknown and must be estimated from the
sample data.

### Types of t-tests:
- One-sample t-test: Compares the sample mean to a specific hypothesized
  population mean. Syntax: `t.test(x, mu = hypothesized_mean)`
- Paired t-test: Compares the means of two paired groups (e.g., before and
  after treatment). Syntax: `t.test(x, y, paired = TRUE)`
- Independent t-test: Compares the means of two independent groups.
  Syntax: `t.test(x, y)  # Assuming equal variances`
  `t.test(x, y, var.equal = FALSE)  # If variances are not equal`

### Confidence Interval
A confidence interval provides a range of values within which the true
population mean is likely to fall with a certain level of confidence
(e.g., 95%). It is calculated using the t-statistic and the sample standard
deviation.

## TESTING A SAMPLE PROPORTION AND ITS CONFIDENCE INTERVAL 
The test of a sample proportion determines if the observed proportion of a
particular characteristic in a sample is significantly different from a
hypothesized population proportion.

## 8.5 COMPARING THE MEANS OF TWO SAMPLES 
Comparing the means of two population samples is a fundamental task in
statistical analysis. To do this in R, the fundamental question to ask relevant
to the use case is: Are you comparing means from independent samples 
(different groups) or paired samples (repeated measures)

### Choosing the appropriate test

##### Independent sample
If the sample is an independent sample, you can use a two sample t-test, a
Welch's t-test or a Wilcoxon rank-sum test.

In R, you can perform a two sample t-test using the `t.test()` with the
appropriate parameters. The two sample t-test assumes the data is normally
distributed. 

A Welch's t.test can be employed if equal variances are unrealistic. In R
you can use the t.test() function but set the paired argument to true like
this `paired = TRUE`

Wilcoxon's rank-sum test is used as a non-parametric alternative for data
that is not normally distributed. The function for this in R is `wilcox.test()` 


##### Paired sample
For normally distributed differences, you can use the `t.test()` with
`paired = TRUE`. Another 

## CODE CHALLENGE
1. Using `table()` create a contingency table named 
   `penguins_contingency_table`  using the species and island columns from the
   penguins dataset.
   
2. Create a contingency table with `xtabs()` and assign to the variable
   `penguins_xtabs_table` using the species and island columns with a formula
   from the Palmer Penguins dataset.
   
3. Calculate the proportions of penguin species for each island using the
   contingency table created in Challenge 1. Assign the resulting proportions
   to a variable named `penguins_species_proportions`
   
4. Conduct an independent two-sample t-test to compare the miles per gallon, of
   cars with 4 cylinders (cyl == 4) and cars with 6 cylinders (cyl == 6) in the
   mtcars dataset. Assign the p-value of the test to a variable named
   `t_test_mpg_vs_cyl_pvalue`

5. Using the penguins dataset, calculate the z-scores for the body mass of
   Chinstrap penguins. Calculate the z-scores relative to the entire dataset
   and assign the resulting vector to a variable named `z_scores_body_mass`

6. Calculate the z-scores for the flipper length (flipper_length_mm) of Adelie
   penguins in the penguins dataset. Calculate the z-scores relative to the
   entire dataset and assign the resulting vector to a variable named
   `z_scores_flipper_length`

7. Use the mtcars dataset. Create a variable named `transmission_table` that
   contains the frequency table of the different transmission types (am column)

8. Using the iris dataset, create a variable named `iris_species_table` that
   contains the frequency table of the different species of iris flowers

9. Use the penguins dataset. Create a cross-tabulation using xtabs() to display
   the counts of penguin species (species) based on their island location.
   Assign the resulting cross-tabulation to a variable named `penguin_cross_tab`

10. Using the electronics data dataset, tabulate how many electronics fell
    under which discount category and assign it to `discount_types`
