# Advanced Topics

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

<script src="files/js/dcl.js"></script>
```{r ,include=FALSE}
tutorial::go_interactive(greedy=TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```

## REPRODUCIBILITY AND REPORTS IN R MARKDOWN

### REPRODUCIBILITY
Reproducibility refers the ability to accurately replicate the results of a 
study or analysis using the same data and methods. It's crucial for scientific
integrity, transparency, and collaboration.

R Markdown is a powerful tool for creating reproducible reports because it
integrates code, text, and output into a single document.

### RMarkdown (.Rmd)
It's a file format that allows you to combine Markdown (a simple text
formatting language) with R code chunks.

You write text, code, and embed results directly in the document. When you
"knit" the document, it's processed to create a formatted output file (HTML,
PDF, Word, etc.), combining text, code, and output seamlessly.

### RMarkdown's Key Features for Reproducibility

1. Code Integration:
  R code chunks are embedded within the document, ensuring all analysis steps
  are documented and executable.
  Readers can easily see the code used to generate results, enhancing
  transparency and understanding.

2. Output Capture:
  The output of R code chunks, such as tables, figures, and statistical
  summaries, is automatically inserted into the final document. This eliminates
  the need for manual copying and pasting, reducing errors and ensuring
  consistency.

3. Version Control
  R Markdown files can be easily versioned using tools like Git, preserving the
  complete history of changes and enabling collaboration.

4. Self-Contained Reports
  The final output document includes both the code and results, making it a
  self-contained and reproducible report.
  Readers can access all necessary information to understand and replicate the
  analysis.

### Using RMarkdown
To create an RMarkdown document, ensure that the rmarkdown package is
installed. You can install it with `install.packages("rmarkdown")`.

Create a new R Markdown document using RStudio or by using the
`rmarkdown::draft()` function.

```{r,tut=TRUE,height=500}
# A simple look of an RMarkdown document

#---
#title: "Reproducible Report"
#author: "Your Name"
#date: "January 1, 2023"
#output: html_document
#---
#
# Introduction
#
#This is an example of an R Markdown document.
#
#```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
#```
```



## RShiny
Shiny is an R package that allows you to build interactive web applications 
directly from R. Shiny makes it easy to create interactive dashboards, data
visualizations, and web applications without the need for web development 
skills. With Shiny, you can turn your R code, analyses, and visualizations into
interactive and user-friendly web applications.

### Key Features of RShiny
- Reactivity: Shiny is built on the concept of reactivity, where changes in
  inputs (e.g., sliders, buttons, text inputs) automatically trigger updates in
  outputs (e.g., plots, tables). This makes it easy to create dynamic and
  responsive applications.

- User Interface (UI): Shiny applications have a user interface that can be
  designed using a combination of R and HTML-like syntax. The UI defines the
  layout and appearance of the application.

- Server Logic: The server logic of a Shiny application is where you define
  the R code that generates outputs based on user inputs. The server function
  is responsible for handling reactive expressions and rendering outputs.

- Widgets: Shiny provides a variety of widgets (input elements) that users can
  interact with, such as sliders, buttons, text inputs, and more. These widgets
  serve as inputs that trigger changes in the application.

- Output Elements: Shiny supports a range of output elements, including 
  plots, tables, and HTML elements. These are updated dynamically in response
  to user interactions.

- Deployment: Shiny applications can be deployed locally or on web servers.
  Popular options for deployment include hosting on shinyapps.io, deploying on
  Shiny Server, or integrating with RStudio Connect.

Here's a simple example of a Shiny application using the mtcars dataset. Run
this in your local RStudio app:

```{r,tut=TRUE,height=500}
# mtcars_shiny_app.R

#library(shiny)
#library(ggplot2)
#
# Load the mtcars dataset
#data(mtcars)
#
# Define UI
#ui <- fluidPage(
#  titlePanel("Scatter Plot of mtcars Variables"),
#  sidebarLayout(
#    sidebarPanel(
#      selectInput("variable", "Select Variable:", choices = colnames(mtcars),
#selected = "disp")
#    ),
#    mainPanel(
#      plotOutput("scatter_plot")
#    )
#  )
#)
#
# Define server logic
#server <- function(input, output) {
#  output$scatter_plot <- renderPlot({
#    # Create a scatter plot based on user-selected variable
#    ggplot(mtcars, aes_string(x = "mpg", y = input$variable)) +
#      geom_point() +
#      labs(title = paste("Scatter Plot of mpg vs.", input$variable),
#           x = "Miles Per Gallon (mpg)",
#           y = input$variable)
#  })
#}

# Create Shiny app
#shinyApp(ui, server)
```

### Some common use cases of RShiny
- Data exploration and visualization
- Interactive dashboards
- Data analysis tools
- Reproducible research
- Educational demonstrations
- Prototypes for web applications


### Why Use RShiny
- Ease of Use: Straightforward syntax and clear structure make it accessible
  to R users.
- Interactivity: Fosters engagement and exploration of data through dynamic
  elements.
- Flexibility: Accommodates a wide range of applications and visualizations.
- Reproducibility: Promotes transparency and sharing of analyses.
- Deployment: Apps can be easily shared online or deployed on servers for wider
  access.



## Building Packages in R
Building packages in R involves creating collections of custom-defined R 
functions, data, and documentation that can be shared and reused by others.

R packages are crucial for sharing code with others, reproducibility, and
maintaining a modular and organized structure for your projects.

R is the or if not, one of the most popular statistical programming languages
and this is in a large part due to its welcome attitude to open-source and its
vibrant community. Many programmers have built R packages which have
significantly expanded the language's applicability across diverse domains,
including healthcare, statistics, education, and research.


### Why Build Packages
- Organization and Reuse: Consolidate code and data into reusable units for
  efficient collaboration and sharing.
- Distribution and Collaboration: Share your work easily with others, fostering
  collaboration and reproducibility.
- Documentation and Maintainability: Improve code quality and clarity through
  documentation and version control.
- Extensibility: Create custom functions and extensions to tailor R to specific
  needs.
- Contribution: Contribute to the R community by sharing your expertise and
  tools.

### Key Steps in Building an R Package
1. Package Structure
  A typical R package has a specific directory structure. The main components
  include the R/ directory for R scripts, man/ for manual pages, data/ for data
  files, inst/ for other files to be installed, and more.
  The DESCRIPTION file contains metadata about the package, including its name,
  version, dependencies, and other information.

2. R Scripts
  Write your R functions and code in the R/ directory. Each function typically
  resides in a separate .R file.
  Document your functions using comments and consider adding documentation
  using the roxygen2 package syntax. Documentation helps users understand how
  to use your functions.

3. Namespace:
  Define a namespace file (NAMESPACE) that explicitly exports functions and
  objects to be accessed by users. This helps avoid conflicts with other
  packages.

4. Documentation
  Use the roxygen2 package to generate documentation from comments in your R
  script files. Documenting your code is essential for helping users understand
  how to use your functions.

5. Data
  If your package includes datasets, place them in the data/ directory. Use the
  data() function in your code to make the data accessible.
  
6. Vignettes
  Vignettes are long-form documentation that provide more extensive
  explanations, use cases, and examples. Place vignettes in the
  vignettes/ directory.
  
7. Tests
  Write test scripts for your functions in the tests/ directory. Testing
  ensures that your functions behave as expected and helps maintain code
  reliability.

8. Build the Package
  Use the devtools package or the built-in R CMD build command to build your
  package. This process creates a .tar.gz file, which is the compressed
  source package.
  
9. Check the Package
  Run R CMD check or use devtools::check() to perform a comprehensive check
  on your package. This ensures that your package complies with best practices
  and does not have any issues.

10. Install and Load
Install your package locally using:
`install.packages("path/to/your/package.tar.gz", repos = NULL, type = "source")`.

Load the package using `library(your_package)`.

11. Version Control
  If you're using version control (e.g., Git), commit your package source files
  and consider sharing your package on platforms like GitHub.

12. Submit to CRAN:
  If you want to share your package with the broader R community, consider
  submitting it to the Comprehensive R Archive Network (CRAN). Follow the CRAN
  submission guidelines.


## END OF COURSE ASSIGNMENT

Everything you learned in this course will be put into test for this take home
project.


You will be using the Wikipedia World Statistics 2023. The final report would
be 2 files, a .Rmd file and a knitted PDF file showing your analysis.

Workflow:
1. Import the dataset into RStudio and ensure you inspect it to see how it
   looks like.
2. Check the data types of each of your variables and make small notes of it.
3. For your analysis, summarize your dataset and using plots, add some
visualization of different variables.
4. Use lapply to convert some of your data types to the appropriate formats
5. Report on the dataset
6. Use linear regression to answert the question, is HDI correlated with IMF's
GDP forecast?
